Towards Verifying the Bitcoin-S Library

### 2019-07-04 paper/paper.tex
more work

\begin{abstract}
  We try to verify properties of the bitcoin-s library, a Scala
  implementation of parts of the Bitcoin protocol. We use the
  Stainless verifier which supports programs in a subset of Scala
  called \emph{Pure Scala}. We first try to verify the property that
  regular transactions do not create new money. It turns out that
  there is too much code involved that lies outside of the supported
  fragment to make this feasible. However, in the process we find and
  fix a bug in bitcoin-s. We then turn to a much simpler property:
  that adding zero satoshis to a given amount of satoshis yields the
  given amount of satoshis. Here as well a significant part of the
  relevant code lies outside of the supported fragment. We perform a
  series of equivalent transformations on it until we arrive at code that we
  successfully verify. In that process we find and fix a second bug in
  bitcoin-s.

\keywords{Bitcoin  \and Scala \and Bitcoin-S \and Stainless.}
\end{abstract}

@@

It comprises algebraic data types in the form of abstract classes,
case classes and case objects, objects for grouping classes and
functions, boolean expressions with short-circuit interpretation,
generics with invariant type parameters, default values of function
parameters, pattern matching, local and anonymous classes and more.

In addition to Pure Scala Stainless also supports some imperative
features, such as using a (mutable) variable in a local scope of a
function and while loops. They turn out not to be relevant for the
current work.

@@

Also, Stainless has its own library of some core data types and
functions which need to be mapped correctly to functions inside of the
SMT solver that Stainless ultimately relies on. Those data types in
general do not have all the methods of the Scala data types. For
example, the BigInt type in Scala has methods for bitwise operations
while the BigInt type in Stainless does not.

@@

\begin{figure}
\begin{lstlisting}[style=scala]
/**
  * Checks the validity of a transaction in accordance to bitcoin core's CheckTransaction function
  * https://github.com/bitcoin/bitcoin/blob/f7a21dae5dbf71d5bc00485215e84e6f2b309d0a/src/main.cpp#L939.
  */
def checkTransaction(transaction: Transaction): Boolean = {
  val inputOutputsNotZero =
    !(transaction.inputs.isEmpty || transaction.outputs.isEmpty)
  val txNotLargerThanBlock = transaction.bytes.size < Consensus.maxBlockSize
  val outputsSpendValidAmountsOfMoney = !transaction.outputs.exists(o =>
    o.value < CurrencyUnits.zero || o.value > Consensus.maxMoney)

@@

  val isValidScriptSigForCoinbaseTx = transaction.isCoinbase match {
    case true =>
      transaction.inputs.head.scriptSignature.asmBytes.size >= 2 &&
        transaction.inputs.head.scriptSignature.asmBytes.size <= 100
    case false =>
      //since this is not a coinbase tx we cannot have any empty previous outs inside of inputs
      !transaction.inputs.exists(_.previousOutput == EmptyTransactionOutPoint)
  }
  inputOutputsNotZero && txNotLargerThanBlock && outputsSpendValidAmountsOfMoney && noDuplicateInputs &&
  allOutputsValidMoneyRange && noDuplicateInputs && isValidScriptSigForCoinbaseTx
}
\end{lstlisting}

@@

A crucial function for the verification of the no-inflation property
is the function which validates transactions: the
\texttt{checkTransaction} function in the ScriptInterpreter object,
shown in Figure~\ref{fig:checktrans}. It takes a transaction and if
some basic checks succeed returns true, otherwise false. For example,
one of those checks is that both the list of inputs and list of
outputs need to be non-empty.

@@

We first create a keypair, then a lock script with its public key,
then the amount of satoshis, then a transaction output (utxo) for that
amount and locked with that script. Finally we create the actual
transaction with that output and no inputs. Of course, that is not a
valid transaction, because it creates coins out of nothing. In
particular, checkTransaction(prevTx) returns false, simply because the
list of inputs is empty.

@@

Now that we have a transaction output, we create a transaction to
spend it.

First, we need a so-called \emph{out point}, which points to an output
of a previous transaction. The second parameter is the index of the
referenced output. Here we use zero because the previous transaction
has only one output (with index zero).

Second, we need some information on how to spend the utxo from the
previous transaction, in particular, how to sign the transaction to
allow this. That is the information needed to create a transaction
input, so essentially a transaction input.

Third, we assemble the list of transaction inputs, in our case just one.

Fourth, we set the amount of satoshis that we want to spend. The Int64
class emulates C data types in Scala, and we will look at it more in
the next section.

Fifth, we create a lock script to send the coins to. To be supremely
secure, we use a public key derived from a private key we immediately
forget.

Sixth, we create our list of transaction outputs.

Seventh and eighth, we define the fee rate and some bitcoin network parameters.

Nineth, we create a transaction builder with those data.

Tenth, we tell the transaction builder to start signing the transaction.

@@

\paragraph{A Bug in the checkTransaction Function.} Note lines 16 and
17 of the checkTransaction function in Figure~\ref{fig:checktrans}.

The value prevOutputTxIds gathers all transaction IDs referenced by
the inputs of the current transaction.  When we call \emph{distinct}
on the returned list, we get the list with duplicates removed.  If the
size of the new list is the same as the size of the old, we know that
there is no duplicate transaction ID.

However, this code prevents a transaction with two inputs spending two
different outputs of a previous transaction. The check is too strict:
checkTransaction returns false for valid transactions.

The fix is simple: we perform the duplicate check on the
TransactionOutPoints instead of on their transaction IDs.  A
TransactionOutPoint contains the txId as well as the output index it
references.

Specifically, we replace lines 16-17 as follows:
\begin{lstlisting}[style=scala, firstnumber=16]
  val prevOutputs = transaction.inputs.map(_.previousOutput)
  val noDuplicateInputs = prevOutputs.distinct.size == prevOutputs.size
\end{lstlisting}

Note that TransactionOutPoint is a case class and thus has a built in
== method.

We submitted the fix together with a unit test to prevent this bug
from appearing again in the future in a pull request
\cite{BitcoinS:pull435}, which has already been merged.

\paragraph{An Attempt at Verification.} Naively trying Stainless on
the entire bitcoin-s codebase results in many errors -- as was to be
expected. We tried to extract only the relevant code and to verify
that. However, even the extracted code has more than 1500 lines and
liberally uses Scala features outside of the supported fragment.  We
tried to transform the code into the supported fragment, but quickly
realized that that a better approach is to first verify a simpler
property with less code involved and later come back to the
no-inflation property with more experience. So we now turn to the
addition-with-zero property.

@@

It is of course a crucial property we are verifying here: if zero
satoshis were credited to your account, you would not want your
balance to change! It is also the simplest meaningful property to
verify that we can think of. However, the code involved in performing
an addition of two satoshi amounts in bitcoin-s is non-trivial. The
reason for that is a pecularity of consensus code: agreement with the
majority is more important than correctness, whatever correctness
might mean. The most widely used bitcoin implementation by far is the
reference implementation Bitcoin Core, written in C++. For consensus
code, bitcoin-s has little choice but to be in strict agreement with
the reference implementation. To achieve that, it implements C-like
data types in Scala and then implements functionality using those
C-like data types. For example, the Satoshis class, which is used to
represent an amount of satoshis, is implemented using the class Int64
which represents the C-type \texttt{int64\_t}.

@@

\paragraph{Extracting the Relevant Code} The relevant code for the
addition of satoshis is in two files: \texttt{CurrencyUnits.scala} and
\texttt{NumberType.scala}. From those files we removed the majority of
the code because it is not needed for the verification of our
property.  For example, we removed all number types except for Int64
(so Int32, UInt64, etc.) because they are not used. We also removed
the superclasses Factory and NetworkElement of CurrencyUnit and
Number, respectively, because the inherited members are not
used. Also, we removed all binary operations on Number that are not
used, like subtraction and multiplication. The extracted code is shown
in Figure~\ref{fig:numbertype} and Figure~\ref{fig:currencyunits}.

\paragraph{A Bug in the checkTransaction Function.}

\paragraph{Transforming the Code.} Here we see how to transform the
bitcoin-s code into the Scala fragment supported by Stainless.  All
transformations are equivalent in the sense that if the
addition-with-zero property holds for the transformed code, then it
also holds for the code before the transformation. Here is the list of
features used by the extracted code but unsupported by Stainless, and
how to transform the code:

\begin{enumerate}
\item Inheriting Objects. Here, we can just turn the objects into case
  objects by literally just changing the word \texttt{object} into
  \texttt{case object} on lines 65 and 33 in the two respective files.

That transformation is equivalent. Case objects have some additional
properties (in particular, being serializable) and they inherit from
\texttt{Product} instead of \texttt{AnyRef}, but none of our code
depends on any of that.

\item Abstract Type Members.

Note that we can not simply replace the unsupported abstract type
member by a (supported) type parameter. The problem is that the
CurrencyUnit class uses one of its implementing classes: Satoshis.

Satoshis would have to instantiate a potential type parameter with
type Int64, so it would extend CurrencyUnit[Int64]. But that is too
specific, because the return type of the \texttt{+}-method would then then be
CurrencyUnit[Int64] not CurrencyUnit[A].

Since we only want to verify the addition of satoshis, and the
Satoshis class overrides A with Int64 anyway, we just remove the
abstract type and set it to Int64.

We remove line 6 and line 22 from CurrencyUnits.scala (to maintain
line numbers we actually replace them with empty lines for now) and in
line 18 we replace A by Int64.

\item{Non-Literal BigInt Constructor Argument}

As described before, the types in the Stainless library are more
restricted than their Scala library counterparts. In particular, the
Stainless BigInt constructor is restricted to literal arguments.

Here we can simply use toBigInt on the field \texttt{underlying}
directly.  So, instead of converting the underlying to Long and back
to BigInt we convert underlying directly to BigInt.

We replace line 26 by
\begin{lstlisting}[style=scala]
def toBigInt: BigInt = underlying.toBigint
\end{lstlisting}

This is an equivalent transformation: the only thing that might go
wrong in the detour via \texttt{Long} is that the BigInt underlying
Int64 in turn underlying Satoshis does not fit into a Long. However,
the only constructor of Int64 ensures exactly that.

\item{Self-Reference in Type Parameter Bound}

Stainless does not currently support a class with a type parameter
with a type boundary that contains the type parameter itself. We opened an
issue \cite{Stainless:issue519} on the Stainless repository and the
Stainless developers have targeted Stainless version 0.4 to support
such self-referential type boundaries.

For now, since our code only uses Number with type parameter T
instantiated to Int64, we just remove the type parameter and replace
it by Int64. We respectively replace lines 8, 44 and 49 by
\begin{lstlisting}[style=scala]
sealed abstract class Number {
\end{lstlisting}

\begin{lstlisting}[style=scala]
sealed abstract class SignedNumber extends Number
\end{lstlisting}

\begin{lstlisting}[style=scala]
sealed abstract class Int64 extends SignedNumber {
\end{lstlisting}

and replace T by Int64 in lines 25-27.

\item{Missing Member \texttt{bigInteger} in BigInt}

The Scala class BigInt is essentially a wrapper around
java.math.BigInteger. BigInt has a member bigInteger which is the
underlying instance of the Java class. The Java class has a method
longValueExact which returns a long only if the BigInteger fits into a
long, otherwise throws exception. Stainless does not support Java
classes and in particular its BigInt has no member bigInteger.

However, our code never calls toLong anymore, so we just remove it. We
replace line 14 in NumberType.scala and line 28 in CurrencyUnits.scala
by an empty line.

\item{Type Member}

Our version of Stainless does not support type members. We just
replace all occurrence of A with BigInt, since A is never overwritten
in an implementing class.

We remove line 9 in NumberType.scala and replace A by BigInt in lines
12, 25, 33 and 50.

In the mean time Stainless has implemented support for type member
\cite{Stainless:pull470}.  Since version 0.2 verification should
succeed without this change.

\item{Missing Bitwise-And Method on BigInt}

Contrary to Scala BigInt, the Stainless BigInt class does not support
bitwise operations, in particular not the \&-method for bitwise and.

The bitwise and with the andMask on line 34 is a bounds check.  It
checks if the result parameter is in range of the specified type,
which in our case is the hard coded Int64.

So, we replace the \& mask with a check whether the result
is in range of Long.MinValue and Long.MaxValue, because Int64 has the
same 64-bit range as Long.

We replace lines 34-35 by the following, squeezed into two lines to
preserve line numbers. Note that the BigInt constructor requires a
literal:

\begin{lstlisting}[style=scala]
require(result <= BigInt("9223372036854775807")  
  && result >= BigInt("-9223372036854775808"),
      "Result was out of bounds, got: " + result)
\end{lstlisting}

\item{Inner Class in Case Object}

Stainless does not support inner classes in a case object. Bitcoin-s
uses this a lot to separate the class from its implementation.

This is easy to fix. We just move the inner classes out of the case objects. They do
not interfere with any other code.

We remove lines 66-71 in NumberType.scala and insert them at the end of the file.
We remove line 42 in CurrencyUnits.scala and insert it at the end of the file.

\item{Message Parameter in Require}

Here the require(condition, message) is used as an assertion: if the
condition is false the fail with message.  Stainless does not support
the message parameter of require.  For the verification we can simply
remove that parameter.

\item{Missing Implicit Long to BigInt Conversion}

This error message does not specify a line number and it is not clear
what ``inv'' is. However, the Scala BigInt has implict conversions
from Long and they are missing in the Stainless BigInt.

We replace the two require clauses at lines 67 and following in NumberType.scala

\begin{lstlisting}[style=scala]
  require(underlying >= -9223372036854775808L)
  require(underlying <= 9223372036854775807L)
\end{lstlisting}

by:
\begin{lstlisting}[style=scala]
  require(underlying >= BigInt(-9223372036854775808L))
  require(underlying <= BigInt(9223372036854775807L))
\end{lstlisting}

We also replace lines 76 and 77
\begin{lstlisting}[style=scala]
  lazy val min = Int64(-9223372036854775808L)
  lazy val max = Int64(9223372036854775807L)
\end{lstlisting}

by:
\begin{lstlisting}[style=scala]
  lazy val min = Int64(BigInt(-9223372036854775808L))
  lazy val max = Int64(BigInt(9223372036854775807L))
\end{lstlisting}

\item{Missing BigInt Constructor with Long Argument}

Here again, the Scala BigInt has a constructor with a Long argument
which is missing in the Stainless BigInt.

We simply replace the Long values in a BigInt constructor call with a string
literal.

The lines from the previous transformation respectively change into
the following:

\begin{lstlisting}[style=scala]
  require(underlying >= BigInt("-9223372036854775808"))
  require(underlying <= BigInt("9223372036854775807"))
\end{lstlisting}

and

\begin{lstlisting}[style=scala]
  lazy val min = Int64(BigInt("-9223372036854775808"))
  lazy val max = Int64(BigInt("9223372036854775807"))
}
\end{lstlisting}
\end{enumerate}

\paragraph{The Specification.}

We can use equals (==) directly on Satoshis, because it is a case class.

The addition will now look like this:
\lstinputlisting[style=scala, firstline=15, lastline=18, firstnumber=15]{../code/addition/src/main/scala/verified/currency/CurrencyUnits.scala}

\paragraph{Result}

Finally, everything is green and correctly verified. 
\begin{figure}
	\centering
		\includegraphics[width=\textwidth]{result_output}
	\caption{Stainless output for the transformed code}
\end{figure}

\section{Conclusion}
\label{chap:conclusion}

Because of the limitations of the verication tool, we could only
verify a rewritten version of the original Bitcoin-S code.  So we can
not guarantee the correctness of the addition of Satoshis with zero in
Bitcoin-S.  Not all changes we made were as trivial as the replacement
of objects with case objects.  For these non-trivial changes, as seen
for example the bound check in section \ref{sec:bound_check}, we
cannot say whether they are equivalent to the original implementation
or not.

So code should be written specically with formal verication in mind,
in order to successfully verify it.  Otherwise, it needs a lot of
changes in the software because verification is mathematical and the
current software is written mostly in object-oriented style.  Software
written in the functional paradigm would be much easier to reason
about.

Thus, either Stainless must find ways to translate more of built-in
object-oriented patterns of Scala to their verification tool or
developers must invest more in functional programming.

@@

\todo{conclusions: what's future work? how to
  change bitcoin-s? how to extend stainless?}

\bibliographystyle{splncs04}
\bibliography{bibliography}

@@

\section{Code Transformations} 
\label{sec:transform}

Here we see in detail how to transform the bitcoin-s code into the
Scala fragment supported by Stainless. All subsections start with the
Stainless error message(s) and finally a description of the changes we
make to the code.

We claim that all transformations are equivalent in the sense that if
the addition-with-zero property holds for the transformed code, then
it also holds for the code before the transformation.

\subsection{Inheriting Objects}

\begin{lstlisting}[style=stainless]
[ Error  ] number/NumberType.scala:65:1: Objects cannot extend classes or implement traits, use a case object instead
          object Int64 extends BaseNumbers[Int64] {
          ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^...
[ Error  ] currency/CurrencyUnits.scala:33:1: Objects cannot extend classes or implement traits, use a case object instead
          object Satoshis extends BaseNumbers[Satoshis] {
          ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^...
\end{lstlisting}

Here, we can just turn the objects into case objects by literally just
changing the word \texttt{object} into \texttt{case object} on lines
65 and 33 in the two respective files.

That transformation is equivalent. Case objects have some additional
properties (in particular, being serializable) and they inherit from
\texttt{Product} instead of \texttt{AnyRef}, but none of our code
depends on any of that.

\subsection{Abstract Type Members}

\begin{lstlisting}[style=stainless]
[ Error  ] currency/CurrencyUnits.scala:6:3: Stainless doesn't support abstract type members
            type A
            ^^^^^^
\end{lstlisting}

Note that we can not simply replace the unsupported abstract type
member by a (supported) type parameter. The problem is that the
CurrencyUnit class uses one of its implementing classes: Satoshis.

Satoshis would have to instantiate a potential type parameter with
type Int64, so it would extend CurrencyUnit[Int64]. But that is too
specific, because the return type of the \texttt{+}-method would then then be
CurrencyUnit[Int64] not CurrencyUnit[A].

Since we only want to verify the addition of satoshis, and the
Satoshis class overrides A with Int64 anyway, we just remove the
abstract type and set it to Int64.

We remove line 6 and line 22 from CurrencyUnits.scala (to maintain
line numbers we actually replace them with empty lines for now) and in
line 18 we replace A by Int64.

\subsection{Non-Literal BigInt Constructor Argument}

\begin{lstlisting}[style=stainless]
[ Error  ] currency/CurrencyUnits.scala:26:33: Only literal arguments are allowed for BigInt.
            def toBigInt: BigInt = BigInt(toLong)
                                          ^^^^^^
\end{lstlisting}

As described before, the types in the Stainless library are more
restricted than their Scala library counterparts. In particular, the
Stainless BigInt constructor is restricted to literal arguments.

Here we can simply use toBigInt on the field \texttt{underlying}
directly.  So, instead of converting the underlying to Long and back
to BigInt we convert underlying directly to BigInt.

We replace line 26 by
\begin{lstlisting}[style=scala]
def toBigInt: BigInt = underlying.toBigint
\end{lstlisting}

This is an equivalent transformation: the only thing that might go
wrong in the detour via \texttt{Long} is that the BigInt underlying
Int64 in turn underlying Satoshis does not fit into a Long. However,
the only constructor of Int64 ensures exactly that.

\subsection{Self-Reference in Type Parameter Bound}

\begin{lstlisting}[style=stainless]
[ Error  ] number/NumberType.scala:8:30: Unknown type parameter type T
          sealed abstract class Number[T <: Number[T]] {
                                        ^^^^^^^^^^^^^^
\end{lstlisting}

Stainless does not currently support a class with a type parameter
with a type boundary that contains the type parameter itself. We opened an
issue \cite{Stainless:issue519} on the Stainless repository and the
Stainless developers have targeted Stainless version 0.4 to support
such self-referential type boundaries.

For now, since our code only uses Number with type parameter T
instantiated to Int64, we just remove the type parameter and replace
it by Int64. We respectively replace lines 8, 44 and 49 by
\begin{lstlisting}[style=scala]
sealed abstract class Number {
\end{lstlisting}

\begin{lstlisting}[style=scala]
sealed abstract class SignedNumber extends Number
\end{lstlisting}

\begin{lstlisting}[style=scala]
sealed abstract class Int64 extends SignedNumber {
\end{lstlisting}

and replace T by Int64 in lines 25-27.

\subsection{Missing Member \texttt{bigInteger} in BigInt}

\begin{lstlisting}[style=stainless]
[ Error  ] number/NumberType.scala:14:22: Unknown call to bigInteger on Number.this.toBigInt (BigInt) with arguments List() of type List()
            def toLong: Long = toBigInt.bigInteger.longValueExact()
                              ^^^^^^^^^^^^^^^^^^^
\end{lstlisting}

The Scala class BigInt is essentially a wrapper around
java.math.BigInteger. BigInt has a member bigInteger which is the
underlying instance of the Java class. The Java class has a method
longValueExact which returns a long only if the BigInteger fits into a
long, otherwise throws exception. Stainless does not support Java
classes and in particular its BigInt has no member bigInteger.

However, our code never calls toLong anymore, so we just remove it. We
replace line 14 in NumberType.scala and line 28 in CurrencyUnits.scala
by an empty line.

\subsection{Type Member}

\begin{lstlisting}[style=stainless]
[Warning ] number/NumberType.scala:9:3: Could not extract tree in class: type A = BigInt (class scala.reflect.internal.Trees$TypeDef)
              type A = BigInt
              ^^^^^^^^^^^^^^^
\end{lstlisting}

Our version of Stainless does not support type members. We just
replace all occurrence of A with BigInt, since A is never overwritten
in an implementing class.

We remove line 9 in NumberType.scala and replace A by BigInt in lines
12, 25, 33 and 50.

In the mean time Stainless has implemented support for type member
\cite{Stainless:pull470}.  Since version 0.2 verification should
succeed without this change.

\subsection{Missing Bitwise-And Method on BigInt}

\begin{lstlisting}[style=stainless]
[ Error  ] number/NumberType.scala:34:14: Unknown call to & on result (BigInt) with arguments List(Number.this.andMask) of type List(BigInt)
              require((result & andMask) == result,
                      ^^^^^^^^^^^^^^^^
\end{lstlisting}

Contrary to Scala BigInt, the Stainless BigInt class does not support
bitwise operations, in particular not the \&-method for bitwise and.

The bitwise and with the andMask on line 34 is a bounds check.  It
checks if the result parameter is in range of the specified type,
which in our case is the hard coded Int64.

So, we replace the \& mask with a check whether the result
is in range of Long.MinValue and Long.MaxValue, because Int64 has the
same 64-bit range as Long.

We replace lines 34-35 by the following, squeezed into two lines to
preserve line numbers. Note that the BigInt constructor requires a
literal:

\begin{lstlisting}[style=scala]
require(result <= BigInt("9223372036854775807")  
  && result >= BigInt("-9223372036854775808"),
      "Result was out of bounds, got: " + result)
\end{lstlisting}

\subsection{Inner Class in Case Object}

>>>>

\begin{abstract}
  We try to verify some properties of the bitcoin-s library, a Scala
  implementation of parts of the Bitcoin protocol. We use the
  Stainless verifier which supports programs in a subset of Scala
  called \emph{Pure Scala}.  Since bitcoin-s is not written in this
  fragment, we extract the relevant code from it and perform a series
  of equivalent transformations until we arrive at code that we
  successfully verify. In that process we find and fix two bugs in
  bitcoin-s.

\keywords{Bitcoin  \and Scala \and bitcoin-s \and Stainless.}
\end{abstract}

@@

It comprises algebraic data types in the form of abstract classes,
case classes and case objects, objects for grouping classes and
functions, boolean expressions with short-circuit interpretation,
generics with invariant type parameters, default values of function
parameters, pattern matching, local and anonymous classes and more.
In addition to Pure Scala Stainless also supports some imperative
features, such as using a (mutable) variable in a local scope of a
function and while loops. They turn out not to be relevant for the
current work.

@@

Also, Stainless has its own library of some core data types and
functions which are mapped to corresponding data types and functions
inside of the SMT solver that Stainless ultimately relies on. Those
data types in general do not have all the methods of the Scala data
types. For example, the BigInt type in Scala has methods for bitwise
operations while the BigInt type in Stainless does not.

@@

\begin{figure}
\begin{lstlisting}[style=scala]
def checkTransaction(transaction: Transaction): Boolean = {
  val inputOutputsNotZero =
    !(transaction.inputs.isEmpty || transaction.outputs.isEmpty)
  val txNotLargerThanBlock = 
    transaction.bytes.size < Consensus.maxBlockSize
  val outputsSpendValidAmountsOfMoney = 
    !transaction.outputs.exists(o =>
      o.value < CurrencyUnits.zero || o.value > Consensus.maxMoney)

@@

  val isValidScriptSigForCoinbaseTx = transaction.isCoinbase match {
    case true =>
      transaction.inputs.head.scriptSignature.asmBytes.size >= 2 &&
        transaction.inputs.head.scriptSignature.asmBytes.size <= 100
    case false =>
      !transaction.inputs.exists(
        _.previousOutput == EmptyTransactionOutPoint)
  }
  inputOutputsNotZero && txNotLargerThanBlock && 
  outputsSpendValidAmountsOfMoney && noDuplicateInputs &&
  allOutputsValidMoneyRange && noDuplicateInputs && 
  isValidScriptSigForCoinbaseTx
}
\end{lstlisting}

@@

A crucial function for the verification of the no-inflation property
is the \texttt{checkTransaction} function shown in
Figure~\ref{fig:checktrans}. Given a transaction it returns true if
some basic checks succeed, otherwise false. For example, one of those
checks is that both the list of inputs and list of outputs need to be
non-empty.

@@

We first create a keypair, then a lock script with its public key,
then the amount of satoshis, then a transaction output (utxo) for that
amount and locked with that script. Finally we create a transaction
with that output and no inputs. Of course, that is not a valid
transaction, because it creates coins out of nothing. In particular,
\texttt{checkTransaction(prevTx)} returns false, simply because the list of
inputs is empty.

@@

Now that we have a transaction output, we create a transaction to
spend it in Figure~\ref{fig:tx}. First, we need a reference to an
output of a previous transaction, here called
\texttt{outPoint}. Second, we add some information on how to spend
that output, in particular, how to sign the transaction to allow
this. Now we assemble the list of unspent transaction outputs (utxos),
in our case just one.

We then set the amount of satoshis that we want to spend. The Int64
class aims to emulate a C data type in Scala, and we will look at it
more closely in the next section.

We then create a lock script (\texttt{destinationSPK}) to receive the
coins, create our list of transaction outputs (\texttt{destinations}),
define the fee rate and set some bitcoin network parameters.

Now we create a transaction builder with those data and we tell it to
start signing the transaction in line 34.

@@

\paragraph{A Bug in the checkTransaction Function.} Note lines 15-17
of the checkTransaction function in Figure~\ref{fig:checktrans}. Here,
value prevOutputTxIds gathers a list of all transaction indentifiers
referenced by the inputs of the current transaction. If the size of
this list is the same as the size of this list with duplicates
removed, we know that no transaction has been referenced twice. This
prevents a transaction from spending two different outputs of the same
previous transaction. The check is too strict: checkTransaction
returns false for valid transactions.

The fix is simple: we perform the duplicate check on the
TransactionOutPoints instead of on their transaction identifiers.  A
TransactionOutPoint contains the txId as well as the output index it
references. Note that TransactionOutPoint is a case class and thus has
a built in == method.

Specifically, we replace lines 15-17 as follows:
\begin{lstlisting}[style=scala, firstnumber=15]
  val prevOutputs = transaction.inputs.map(_.previousOutput)
  val noDuplicateInputs = 
    prevOutputs.distinct.size == prevOutputs.size
\end{lstlisting}

We submitted this fix together with a corresponding unit test to the
bitcoin-s project in a pull request, which has been merged
\cite{BitcoinS:pull435}.

\paragraph{An Attempt at Verification.} Naively trying Stainless on
the entire bitcoin-s codebase results in many errors -- as was to be
expected. We tried to extract only the code relevant to the
no-inflation-property and to verify that. However, even the extracted
code has more than 1500 lines and liberally uses Scala features
outside of the supported fragment.  We tried to transform the code
into the supported fragment, but quickly realized that a better
approach is to first verify a simpler property with less code involved
and later come back to the no-inflation property with more
experience. So we now turn to the addition-with-zero property.

@@

It is of course a crucial property we are verifying here: if zero
satoshis were credited to your account, you would not want your
balance to change! It is also the simplest meaningful property to
verify that we can think of. However, the code involved in performing
the addition of two satoshi amounts in bitcoin-s is non-trivial. The
reason for that is a pecularity of consensus code: agreement with the
majority is more important than correctness, whatever correctness
might mean. The most widely used bitcoin implementation by far is the
reference implementation Bitcoin Core, written in C++. For consensus
code, bitcoin-s has little choice but to be in strict agreement with
the reference implementation. To achieve that, it implements C-like
data types in Scala and then implements functionality using those
C-like data types. For example, the Satoshis class, which represents
an amount of satoshis, is implemented using the class Int64 which
aims to represent the C-type \texttt{int64\_t}.

@@

\paragraph{Extracting the Relevant Code} The relevant code for the
addition of satoshis is in two files: CurrencyUnits.scala and
NumberType.scala. From those files we removed the majority of the code
because it is not needed for the verification of our property.  For
example, we removed all number types except for Int64 (so Int32,
UInt64, etc.) because they are not used. We also removed the
superclasses Factory and NetworkElement of CurrencyUnit and Number,
respectively, because the inherited members are not used. Also, we
removed all binary operations on Number that are not used, like
subtraction and multiplication. The extracted code is shown in
Figure~\ref{fig:numbertype} and Figure~\ref{fig:currencyunits}.

\paragraph{A Bug in the checkResult Function.} Note the checkResult
function on line 12 and the value andMask on line 23 of
NumberType.scala. The function is intended to catch overflows by
performing a bitwise conjunction of its argument with andMask and
comparing the result with the argument. However, because of the way
Java BigIntegers are represented \cite{wikipedia:twocomp} and because
bitwise operations implicitly perform a sign extension
\cite{java:bigint} on the shorter operand, the function does not
actually catch overflows.

While this is a potentially serious bug, it turns out that checkResult
is only ever called inside a constructor call for a number type (like
Int64) which contains the intended range check, see lines 32-35. The
checkResult function thus can, and should, be removed entirely. The
bitcoin-s developers have acknowledged the bug and we submitted a pull
request to fix it \cite{BitcoinS:pull565}.

\paragraph{Transforming the Code.} We now turn to the list of Scala
features used by the extracted code which are not supported by
Stainless and how to transform the code into the supported fragment.
All transformations are equivalent in the sense that if the
addition-with-zero property holds for the transformed code, then it
also holds for the code before the transformation.

\paragraph{Inheriting Objects.} In both files we have objects
extending the BaseNumbers trait, on lines 30 and 23 respectively,
which Stainless does not support. We simply turn those onjects into
case objects. That transformation is equivalent. Case objects have
various additional properties (for example, being serializable) but
none of our code depends on the absence of any of that.

\paragraph{Abstract Type Members.} In CurrencyUnits.scala on line 6
there is an abstract type that is not supported. Note that we can not
simply replace it by a (supported) type parameter since the
CurrencyUnit class uses one of its implementing classes:
Satoshis. Since the Satoshis class overrides A with Int64 anyway, we
just remove the abstract type declaration and replace A by Int64
everywhere.

\paragraph{Non-Literal BigInt Constructor Argument.} In
CurrencyUnits.scala on line 18 the BigInt constructor is called with a
non-literal argument. As described before, the types in the Stainless
library are more restricted than their Scala library counterparts. In
particular, the Stainless BigInt constructor is restricted to literal
arguments. So we simply replace \texttt{toLong} by
\texttt{underlying.toBigInt}: instead of converting the underlying
Int64 (which in turn has an underlying BigInt) to Long and then back
to BigInt we simply directly return the BigInt. This is an equivalent
transformation: the only thing that might go wrong in the detour via
\texttt{Long} is that the underlying BigInt does not fit into a
Long. However, the only constructor of Int64Impl ensures exactly that
and all functions producing Int64 do so via this constructor.

\paragraph{Self-Reference in Type Parameter Bound.} In
NumberTypes.scala on lines 3 and 19 are classes a type parameter and a
type boundary that contains that typarameter itself. Stainless does
not currently support such self-referential type boundaries. We opened
an issue \cite{Stainless:issue519} on the Stainless repository and the
developers have targeted version 0.4 to support self-referential type
boundaries. Since our code only uses Number with type parameter T
instantiated to Int64, we just remove the type parameter declaration
and replace all its occurrences it by Int64.

\paragraph{Missing Member \texttt{bigInteger} in BigInt.} In NumberType
on line 6 there is a reference to \texttt{bigInteger}. The Scala
BigInt class is essentially a wrapper around
java.math.BigInteger. BigInt has a member bigInteger which is the
underlying instance of the Java class. The Java class has a method
longValueExact which returns a long only if the BigInteger fits into a
long, otherwise throws exception. Stainless does not support Java
classes and in particular its BigInt has no member
bigInteger. However, our code never calls toLong anymore, so we just
remove it.

\paragraph{Type Members.} In NumberType.scala there is a type member
on line 4. Our version of Stainless (0.1) does not support type
members. We just remove the declaration and replace all occurrences of
A with BigInt, since A is never overwritten in an implementing class.
Note that in the mean time Stainless has implemented support for type
members \cite{Stainless:pull470}.  Since version 0.2 verification
should succeed without this change.

\paragraph{Missing Bitwise-And Method on BigInt.} Contrary to Scala
BigInt, the Stainless BigInt class does not support bitwise
operations, in particular not the \&-method used in NumberType.scala
on line 13. However, as described above, the checkResult function is
both broken and redundant, so we remove it and all calls to it.

\paragraph{Inner Class in Case Object.} We have inner classes in
NumberType.scala on line 31 and in CurrencyUnits.scala on line
26. Stainless does not support inner classes in a case object. We just
move the inner classes out of the case objects. They do not interfere
with any other code.

\paragraph{Message Parameter in Require.} The calls of the require
function on lines 32 and 34 in CurrencyUnits.scala have a second
parameter: the error message. Stainless does not support the message
parameter. We simply remove it.

\paragraph{Missing Implicit Long to BigInt Conversion.} The Scala
BigInt class has implict conversions from Long which NumberType.scala
uses on lines 32 and 34 and they are missing in the Stainless
BigInt. Since a BigInt constructor with a Long argument is also
missing, we replace the Long literals by an explicit call to the
BigInt constructor with a literal string argument,
e.g. \texttt{BigInt("-9223...5808")}.

\paragraph{The Specification.} Now that all our code has been
transformed into the supported fragment, we can finally write our
specification, shown in Figure~\ref{fig:spec}, and verify it with
Stainless, as the output in Figure~\ref{fig:result} shows.

\begin{figure}
\lstinputlisting[style=scala, firstline=9, lastline=12, firstnumber=9]{../code/addition/src/main/scala/verified/currency/CurrencyUnits.scala}
\caption{Addition function with specification}
\label{fig:spec}
\end{figure}

\begin{figure}
	\centering
		\includegraphics[width=\textwidth]{result_output}
	\caption{Stainless output for the transformed code}
  \label{fig:result}
\end{figure}

\section{Conclusion and Future Work}

We have show
-- bugs (already fixed)
-- suggestions for bitcoin-s

Because of the limitations of the verication tool, we could only
verify a rewritten version of the original Bitcoin-S code.

So code should be written specically with formal verication in mind,

@@

\bibliographystyle{splncs04}

\bibliography{bibliography}

@@
